Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-14 writeup/conStruct.tex
merged P's changes and added Fig4

\title{Inferring Continuous and Discrete Population Genetic Structure Across Space}
\date{\vspace{-5ex}}
\author{
Gideon S.\ Bradburd$^{1,a}$, 
Graham M.\ Coop$^{2,b}$,
Peter L.\ Ralph$^{3,c}$}

@@

\plr{I felt like we said "continuous and discrete" too many times in the abstract.}

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species, because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
Descriptions of taxa using phylogenetic trees omit within-taxon variation,
especially when these are intermingling subgroups.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation,
\plr{as for example XXX?}
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

@@

This definition of covariance differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if these are likely to contain a large percentage of errors.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions by which individuals' ancestry derive from each layer 
(the ``admixture'' proportions) 
do so as well.

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the cluster.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
%% plr: this is no longer true
% The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
% connecting the $k$th population to the population ancestral to all modeled
% layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats}. 
In the methods section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} in the appendix for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in computational speed.)
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our non-spatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} of the appendix for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

Another way to describe the practical significance of additional layers
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers if they are present, 
so that they will have a low contribution to overall covariance.
To do this, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some background probability,
but this probability of agreement is increased on any segment of genome 
that both have inherited through the same layer.
(The amount it increases depends on how far apart they are.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

\subsection*{Simulations}

To test the method, we first generated data using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture, spatially autocorrelated proportions 
(see Appendix \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers,
in each sampling 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation to compare the predictive performance.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, using $K=2--4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

\plr{rearranged this to focus more on take-homes}

% In Fig \ref{simK1_pies}, we show the results of the dataset simulated with $K=1$, 
% contrasting the results of the nonspatial model runs with 2--4 layers 
% with those of spatial model with 2--4 layers.
% 
% In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
% The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
% Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
% 
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
% we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

With these simulations,
spatial \texttt{conStruct} does not create spurious clusters when there are none:
Figures \ref{simK1_pies}, \ref{simK2_sp_pies}, and \ref{simK3_sp_pies} 
show that subsequent layers beyond the number used for simulation are unused 
when data is simulated with $K=1$ but analyzed with $K>1$ --
the layers are present, but contribute very little to any population.
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

In contrast, the nonspatial model describes geographic variation
using gradients of admixture between more and more discrete clusters
to better approximate the continuous, spatial patterns of relatedness
(depicted in Figs \ref{simK1_pies} and \ref{simK1_nsp_pies}).
% as $K$ in the model increases, 
% the sampled area is increasingly divvied up between clusters 
% with ``centers of mass" in the $K$ geographical extremes of the simulated space.
% At $K=2$, the clusters are centered in opposite corners on the map; 
% at $K=3$, in the lower two corners and the top half; 
% at $K=4$, all four corners, and so on.
The fastSTRUCTURE results are qualitatively similar, as shown in Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}.
Each non-spatial cluster is in truth genetically more similar within itself
than it is to other clusters,
but we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}),
while the predictive accuracy of the nonspatial model increases as subsequent clusters are added.

% In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
% either plateaus at or declines from the true value of $K$ used to simulate the data 
% (Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).

The unimportance of spurious layers can be seen in plots of layer contributions,
Figs \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}.
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional clusters in the nonspatial analyses continue to contribute substantially.

@@

\subsection*{Empirical Applications}

To demonstrate the utility of this method, we analyzed empirical population genomic data from two systems:
a contact zone between two poplar species in northwestern North America,
and a large North American sample of black bears.

@@

\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus balsamifera}, the balsam poplar,
have a broad zone of overlap in the Pacific Northwest,
where they are hypothesized to hybridize \citep{geraldes_etal_2014, suarezgonzalez_etal_2016}.
Both species are sampled over a large geographic region, 
and show spatial patterns of genetic and phenotypic variation \citep{slavov_etal_2012, mckown_etal_2013},
making the system well-suited for application of our method.
% to test an analysis designed to describe
% how genetic structure is partitioned both discretely across populations (or species)
% and continuously with geographic distance.
We organize the results of our analyses around the following questions: 
\begin{enumerate}

    \item To what degree has hybridization blurred the boundaries between 
        \textit{trichocarpa} and \textit{balsamifera}?
        (As an extreme case, does genetic differentiation support these as separate species,
        as opposed to a single cline of ancestry?)
    % Given the known history of hybridization between this pair of poplar species, 
    % is there support for two discrete populations, 
    % rather than a single cline of ancestry across their overlap zone?

    \item Does the only significant boundary of population structure
        fall along the species boundary (if any),
        % If there is evidence for discrete population structuring, 
        % does it fall solely along species boundaries, 
        or is there sub-structuring within species?

    \item Does the strength of isolation by distance differ between inferred layers?
        This may indicate, e.g., different speeds of postglacial expansion
        or primary modes of dispersal.
        % Is there support for a pattern of isolation by distance within
        % the inferred layers, and, if so, 
        % does it differ between different layers?

\end{enumerate}

We use data from \cite{geraldes_etal_2014}, 
consisting of 434 individuals sampled from 35 drainages 
genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
\plr{How do the indivs partition across the drainages?  Perhaps give range of number of indivs per drainage?}
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
For comparison, we also ran fastSTRUCTURE \citep{fastStructure} 
using $K = 1$ through 7.
The results of all these analyses are shown in 
Figs \ref{populus_pies} and \ref{populus_xvals}, 
as well as in Figs \ref{populus_sp_pies} - \ref{populus_fastStr} in the Supplementary Materials.

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=2in]{figs/populus/Fig4_pop_sp_results.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{Populus} dataset 
    using the spatial \texttt{conStruct} model for $K=2$ through 4.
    }\label{populus_pies}
\end{figure}

@@

All models with $K>1$ assigned the majority of each of the two species
to distinct layers, with some populations drawing ancestry from multiple layers.
Based on cross-validation results, we view the $K=3$ spatial model 
as a sufficient description of the data,
with additional structure of uncertain significance.
This provides strong support for discrete population structure between the two species,
with some admixture,
rather than a single, continuous cline of ancestry.
At all values of $K>1$, discrete population structure was mostly partitioned along species lines; 
at values of $K$ above 2, further discrete substructure was inferred within the \textit{P. trichocarpa} samples, 
with no substructure within \bals{}.
Most of this signal seems to derive from the \textit{P. trichocarpa} samples:
% covariance within predominantly \tri{} layers decays more quickly with distance than within the majority \bals{} layer, 
as seen in Fig \ref{populus_sp_layer_covs},
there is almost no isolation by distance within the \bals{} layer ($\alpha_D \approx 0$).
Both points are
in agreement with \citet{keller_etal_2010}, 
who found low diversity within the region's \bals{},
probably as the result of a recent postglacial expansion.

For all values of $K$, the spatial model was preferred over the nonspatial model (Fig \ref{populus_xvals}) (strongly for $K<4$).
The nonspatial results were broadly concordant with those of the fastSTRUCTURE runs (Fig \ref{populus_fastStr}).
\gb{come back to this}.

A consistent split between layers within \tri{} fell along the ``no-cottonwood belt," 
a region along the central coast of British Columbia in which black
cottonwood is absent (the break between yellow and red, for $K \geq 3$). 
The no-cottonwood belt is hypothesized to divide the species' distribution
into northern and southern groups, which, in a provenance test, 
were experimentally shown to display differences in ecologically relevant phenotypes 
\citep[e.g., pathogen resistance,][]{xie2009,xie2012}.  
At higher values of $K$, drainages at the southern tip of \tri{} sampling 
begin to split out into their own layers,
perhaps due to introgression from the southern neighbors
\textit{P. angustifolia} or \textit{fremontii} \citep{Zhou2012,geraldes_etal_2014}

\subsubsection*{Black bears}

>>>>

\title{Inferring Continuous and Discrete Population Genetic Structure Across Space}
\date{\vspace{-5ex}}
\author{
Gideon S. Bradburd$^{1,a}$, 
Graham M. Coop$^{2,b}$,
Peter L. Ralph$^{3,c}$}

@@

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, reasonable people can disagree about whether two 
populations are separate species, because the 
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
The issue of identifying meaningful biological subunits 
extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long-distance migration. 
Both discrete and continuous components
are required to accurately describe most species' patterns of genetic relatedness.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

@@

This definition of covariance differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer, allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions of ancestry individuals derive from each layer 
(the ``admixture'' proportions) 
do so as well.

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats},
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In the Appendix section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} in the Appendix for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in computational speed.)
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our non-spatial method 
are a close match to those of STRUCTURE-based approaches.

@@

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} in the Appendix for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

Another way to describe how additional layers contribute to model fit
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers, 
so that they will have a low contribution to overall covariance.
We use the connection of allelic covariance to the probability of agreement (identity) between genotypes
to calculate a measure of overall contribution of each layer to the model.

The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The contribution of the $k$th layer is its absolute contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this contribution, see Appendix \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

\subsection*{Simulations}
We first test the method with scenarios generated using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
To generate discrete population structure, we simulate $K$ distinct populations,
each of which split from a common ancestor $\tau_{\text{s}}$ units of coalescent time in the past,
without subsequent migration between them.
To generate continuous differentiation within each layer,
at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonizes an
independent lattice of demes,
for which we use a stepping stone model with symmetric migration 
to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past 
we generate a single dataset 
by collapsing those $K$ discrete layers into a single grid of demes 
that have various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapse the layers together using random admixture proportions 
that are spatially autocorrelated (see Appendix \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers,
in each sampling 10,000 unlinked loci from each of 20 haploid individuals from every deme.

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the analyses of the dataset simulated under $K=1$, 
contrasting the results of the nonspatial model runs with 2 - 4 layers 
with those of spatial model with 2 - 4 layers.
In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
To confirm that results from our nonspatial model were qualitatively similar to those of 
other nonspatial model-based clustering algorithms, 
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}).

A number of results are immediately clear from these tests with simulated data.
First, \texttt{conStruct} performs well on all simulated scenarios.
Across all values of K and for all simulations, 
the spatial model is correctly preferred over the nonspatial model
(Fig \ref{sim_xvals}). Second, the predictive accuracy of the nonspatial model increases 
with the number of layers in the model, 
as it describes continuous, spatial patterns of relatedness between demes within the same cluster 
using gradients of admixture across more and more discrete clusters.
This behavior of the nonspatial model is highlighted in 
the analyses of data simulated under $K=1$ (Fig \ref{simK1_pies} and \ref{simK1_nsp_pies}):
as $K$ in the model increases, 
the sampled area is increasingly divvied up between clusters 
with ``centers of mass" in the $K$ geographical extremes of the simulated space.
At $K=2$, the clusters are centered in opposite corners on the map; 
at $K=3$, in the lower two corners and the top half; 
at $K=4$, all four corners, and so on.
These results are qualitatively similar to those generated using fastSTRUCTURE \citep{fastStructure}, 
which also describes continuous differentiation using admixture between clusters
 (Fig \ref{fastStr_simK1}-\ref{fastStr_simK3}).
While it is true that each non-spatial cluster is genetically more similar within itself
than it is to other clusters,
we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
either plateaus at or declines from the true value of $K$ used to simulate the data 
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).
And, as $K$ in the spatial model increases, 
deme membership in layers beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
Even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
Moreover, the method infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).

Finally, we calculated layer contribution for 
analyses run in each simulation 
(Fig \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}).
In the spatial analyses, once we pass the true $K$, 
subsequent layers add little in terms of (co)variance explained; 
in contrast, additional clusters in the nonspatial analyses continue to
be important for describing variation, 
as they are forced to partition up the spatial patterns within clusters.

@@

\subsection*{Empirical Applications}
To demonstrate the utility of this method, we analyzed empirical population genomic data from two systems:
a contact zone between two poplar species in northwestern North America,
and a large North American sample of black bears.

@@

Here, we focus on two \textit{Populus} species in the Pacific Northwest: 
\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus balsamifera}, the balsam poplar.
These species have a broad zone of overlap and are hypothesized to hybridize \citep{geraldes_etal_2014, suarezgonzalez_etal_2016}.
Both species are sampled over a large geographic region, 
and show spatial patterns of genetic and phenotypic variation \citep{slavov_etal_2012, mckown_etal_2013},
making the system well-suited to test an analysis designed to describe
how genetic structure is partitioned both discretely across populations (or species)
and continuously with geographic distance.
We organize the results of our analyses around the following questions: 
\begin{enumerate}
\item Given the known history of hybridization between this pair of poplar species, 
is there support for two discrete populations, 
rather than a single cline of ancestry across their overlap zone?
\item If there is evidence for discrete population structuring, 
does it fall solely along species boundaries, 
or is there sub-structuring between different populations of the same species?
\item Is there support for a pattern of isolation by distance within
the inferred layers, and, if so, 
does it differ between different layers?
\end{enumerate}

Our data, from \cite{geraldes_etal_2014}, 
consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
For comparison, we also ran fastSTRUCTURE \citep{fastStructure} 
using $K = 1$ through 7.
The results of all these analyses are shown in 
Figs \ref{populus_pies} and \ref{populus_xvals}, 
as well as in Figs \ref{populus_sp_pies} - \ref{populus_fastStr} in the Supplementary Materials.

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=4in]{figs/populus/Fig4_pop_sp_results.pdf}}
	\caption{
	Maps of admixture proportions estimated for the \textit{Populus} dataset 
	using the spatial model for $K=2$ through 4 (a-c), 
	as well as the corresponding layer-specific covariance curves estimated under each model (d-f).
    }\label{populus_pies}
\end{figure}

@@

All models with $K>1$ assigned the majority of each of the two species
to distinct layers, with some populations drawing ancestry from multiple layers.
This provides strong support for discrete population structure between the two species,
with some admixture,
rather than a single, continuous cline of ancestry.
At all values of $K>1$, discrete population structure was mostly partitioned along species lines; 
at values of $K$ above 2, further discrete substructure was inferred within the \textit{P. trichocarpa} samples, 
while \bals{} samples displayed no substructure up to $K=7$.
This lack of substructure within the sampled \bals{} populations is largely in agreement with results from 
\citet{keller_etal_2010}, who found that \bals{} in the region is characterized by relatively low diversity, 
and probably represents a recent, postglacial expansion.

For values of $K<4$, the spatial model was strongly preferred over the nonspatial model (Fig \ref{populus_xvals}).
Most of this signal seems to derive from the \textit{P. trichocarpa} samples:
covariance within predominantly \tri{} layers decays more quickly with distance
than within the majority \bals{} layer, as seen in Fig \ref{populus_sp_layer_covs}.  
This indicates that, although isolation by distance is a feature of the dataset as a whole, 
its magnitude varies considerably across biological units - 
an aspect of the data \texttt{conStruct} is able capture well.
Based on our cross-validation results we view the $K=3$ spatial model 
as a sufficient description of the data.

A consistent split between layers within \tri{} fell along the ``no-cottonwood belt," 
a region along the central coast of British Columbia in which black
cottonwood is absent (the break between yellow and red, for $K \geq 3$). 
The no-cottonwood belt is hypothesized to divide the species' distribution 
into northern and southern groups, which, in a provenance test, 
were experimentally shown to display differences in ecologically relevant phenotypes 
\citep[e.g., pathogen resistance,][]{xie2009,xie2012}.  
At higher values of $K$, drainages at the southern tip of \tri{} sampling 
begin to split out into their own layers.  
This may potentially be the result of introgression from southern \textit{Populus} species, 
such as \textit{P. angustifolia} or \textit{fremontii} \citep{Zhou2012,geraldes_etal_2014}

\gb{The results of the nonspatial analysis differed from those of the fastSTRUCTURE runs (Fig \ref{populus_fastStr}), 
although both showcased the tendency of nonspatial models to describe continuous variation with multiple clusters.
Replicate runs of the nonspatial \texttt{conStruct} model produced different results, 
evidence of multimodality in the likelihood surface, 
which we might expect given the poor correspondence of the model to the data.}
